Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2014-11-26 Rmarkdown/wgbs_analysis.Rmd
Split WGBS analysis into “bioinformatics” and “statistics”

---
title: "Bioinformatics analysis of bisulfite-sequencing data"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

@@

This chapter outlines the steps in a bioinformatics analysis of a whole-genome bisulfite-sequencing dataset, with a focus on experiments studying _differential methylation_. Throughout this chapter I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay. All data used in my thesis were generated using this protocol (__CHECK__). Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some additional tweaks.

There are four fundamental steps in the bioinformatics analysis:

1. Data quality control checks
2. Read mapping and post-processing of mapped reads
3. Methylation calling
4. Identifying differential methylation

__FIGURE__ provides a more detailed view of these four steps (__Figure should be a flow chart with the various sub-steps included, e.g. Inference = DMC, DMR, ASM, meth-SNP detection, etc.__)

Steps 1, 2 and 4 will be familiar to anyone who analyses high-thoughput sequencing data, but each requires a "twist" to work with bisulfite-sequencing data. Step 3 is obviously unique to assays of DNA methylation but there are similarities to variant calling from DNA-seq.

In this chapter I describe steps 1-2 in broad terms. An excellent review of these fundamental steps, including comparisons of different software, is given by \citet{Krueger:2012ks}. My thesis has focused on steps 3 and 4 of this process. In this chapter I review previous work on steps 3-4 and defer my contributions on these topics to subsequent chapters. While this chapter focuses on analysing bisulfite-sequencing data, in particular whole-genome bisulfite-sequencing data, some parts __SECTION Cellular heterogeneity, Inference, OTHERS???__ will also be applicable to methylation microarrays (__CHECK AFTER I'VE WRITTEN IT__).

@@

The `FastQC` software (__CITE__) is a very useful tool for performing this first step. It produces summary graphs of many key measures such as base quality scores, read length distribution and sequence contamination. `FastQC` is a general purpose tool for performing quality control checks of high-throughput sequencing data. This means that some of its output is not so useful, or at least must be cautiously interpreted, for bisulfite-sequencing data. For example, `FastQC` will report a __warning/error__ if the GC-percentage, the percentage of read bases that are guanine or cytosine, is __less than what value__. Due to the bisulfite-treatment, bisulfite-sequencing data will naturally have a very low percentage of cytosine bases (__around what%__) and therefore a low GC-percentage. Therefore, a low GC-percentage is no cause for concern.

More important is the identification and removal of contaminating sequences. FastQC will screen a subset of the reads against a list of known, common contaminants. Examples of such contaminants include the sequencing adapters, PCR primers and __another example__.

Illumina adapter sequences are ligated to each DNA molecule in the library in order to perform Illumina sequencing. The sequencer can "read into" the adapter sequence, particularly when using paired-end sequencing of short DNA fragments such as those created in WGBS libraries. This means that some reads are a chimera containing sequence of interest (from the sample) and junk (from the adapters). This junk needs to be removed for two reasons:

1. Reads containing adapter contamination will generally not map to the reference genome, meaning these reads go to waste.
2. If they do map, then this will result incorrect inferences - the garbage in, garbage out maxim.

Using a tool such as `Trim Galore!` (__CITE__) or `trimmomatic` (__CITE__), the reads can be _trimmed_ to remove these contaminants. Reads might also be trimmed to remove low quality positions, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.

@@

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are now mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed.

@@

1. "Methylation-aware" mismatch penalties.
2. _In silico_ bisulfite-conversion of reads and reference genomes.

While _methylation-aware_ mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads (__CITE FELIX__). This biases downstream inference and means that these mappers are generally less popular. I will instead focus on the _in silico_ bisulfite-conversion mappers, called as such because of a key step taken by these in the mapping process.

_In silico_ bisulfite-conversion mappers convert all cytosines to thymines (resp. guanines to adenines) of the forward (resp. reverse) strand from the reference genome. They then take each read and create two _in silico_ bisulfite-converted versions of it[^2_reads]: the CT-read replaces all residual thymines with cytosines and the GA-read replaces all residual guanines with adenines. The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as Bowtie2 (__CITE__) or BWA (__CITE__). This process is illustrated in __FIGURE__.

@@

Depending on the exact settings used, the mapper reports the "best" location of each read with respect to the two reference genomes. It reports the original sequence of the read in the output file so that the methylation status of each position can be inferred by comparing it to the corresponding reference sequence.

_In silico_ bisulfite-conversion mappers avoid the bias inherent in the "methylation-aware" mappers because all reads, regardless of methylation status, "look the same" to the mapper. However, they do suffer from a slight loss in mapping efficiency (__CITE FELIX__).

__TABLE__ lists some popular bisulfite-sequencing read mappers with the underlying mapping software listed in brackets:

* Bismark (Bowtie1 or Bowtie2, __CITE__)
* BWA-meth (BWA, __CITE__)
* BSMAP (SOAP, __CITE__)
* __OTHERS__

Each of these aligners can report the output in the standard `SAM` format (__CITE__), although each mapper does so in a slightly different way, which it difficult to develop downstream analysis tools.

@@

The above problems are general challenges of read mapping and are not specific to bisulfite-sequencing data, although the reduced complexity of bisulfite-sequencing reads exacerbates these issues. The difficulty of mapping to repetitive regions of the genome is a particularly frustrating one for  bisulfite-sequencing data. Repetitive sequences, such as LINEs and SINEs, are frequently methylated and so are of interest to researchers studying DNA methylation. The low mapping efficiencies of these regions means that there is often limited data of these elements from bisulfite-sequencing data (__CHECK IN MY DATA AND REPORT COVERAGE__).

@@

PCR amplification of the input DNA is a common step in creating a library for high-throughput sequencing. PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work. Unfortunately, it can introduce biases into the library that result in some molecules being over- or under-represented compared to their "true" frequency. This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome. These multiply-sequenced fragments are called _PCR duplicates_.

In bisulfite-sequencing data, PCR duplicates containing methylation locus $i$ result in $\beta_{i}$ being a biased estimate of $B_{i}$. This is because the sequenced reads do not accurately represent the true methylation levels of the sample.

Generally, there is no way to tell based on sequencing data if a read is truly a PCR duplicate. However, it is relatively easy to identify __possible__ PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]). Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` library (__CITE__), the `rmdup` function that is a part of the `SAMtools` library (__CITE__) and `SAMBLASTER` \citep{Faust:2014hf}. These software use mapped reads from a `SAM/BAM` file as input. There are subtle differences in how each method calls PCR duplicates. As an example, this is how `Picard`'s `MarkDuplicates` function works:

[^pcr_dup]: The distinction between possible PCR duplicates and true PCR duplicates is rarely made, probably because the phrase is so clunky. Possible PCR duplicates are almost always referred to as PCR duplicates with the implicit assumption that the reader is aware that these very likely include falst positive calls. Consistent with the literature, I will use the term PCR duplicates when I refer to reads identified as PCR duplicates by some software. I will use _true_ PCR duplicates when I need to distinguish the two concepts.

> Essentially what it does (for pairs; single-end data is also handled) is to find the 5' coordinates and mapping orientations of each read pair. When doing this it takes into account all clipping that has taking place as well as any gaps or jumps in the alignment. You can thus think of it as determining "if all the bases from the read were aligned, where would the 5' most base have been aligned". It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair. "Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15 (Source: [http://sourceforge.net/apps/mediawiki/picard/index.php?title=Main_Page#Q:_How_does_MarkDuplicates_work.3F](http://sourceforge.net/apps/mediawiki/picard/index.php?title=Main_Page#Q:_How_does_MarkDuplicates_work.3F)).

Using these software will result in false positive calls because reads may satisfy this (or similar) criteria yet come from separate DNA fragments in the original sample. The false positive rate is a particular problem when a small "genome", which might be a targetted region of a genome such as exons or CGIs, is sequenced at high coverage. This can be thought of as an example of the pigeonhole princple, which states that if we have $m$ containers (positions where a read can start) and $n > m$ items (reads), then at least one container must contain more than one item (at least one position must have more than one read starting there).

We could make this mathematically more precise, but it doesn't give us a simple answer to the question, "should we remove possible PCR duplicates from bisulfite-sequencing data?". The unsatisfactory answer is, "it depends". A rough guide is that provided the average/median sequencing coverage of the "genome" is less than the fragment length[^pcr_fragment_length] then we expect few false positive calls.

[^pcr_fragment_length]: For single-end data, the "fragment length" in these calculations is the read length.

@@

Ideally, the probability that a base is called methylated should be independent of the sequencing cycle. \citet{Hansen:2011gu} found that this is not the case and that in fact there is considerable bias towards the start (5') and end (3') of reads. They called this bias the _M-bias_.

M-bias can be identified by plotting the read-position methylation level ($rpml$), which is the proportion of reads that are methylated at each read-position, as a function of read-position (see __FIGURE__). These $rpml$ are computed separately for each methylation type and, for paired-end sequencing data, separately for `read_1` and `read_2`. If there is no M-bias then this plot should be a horizontal line. A "bend" or "spike" in this plot is evidence of M-bias. Furthermore, these lines, which indicate the average level of methylation in the sample for that methylation type, should be at the same height for `read_1` and `read_2`, although we see this is often not quite the case (see __FIGURE__).

If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should perhaps be performed separately for each batch and then combined, or in some manner that is "batch aware". For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs. Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with.

The strongest source of M-bias in Illumina WGBS data is at the 5' end of `read_2`, which sequences the 3' end of the DNA fragment. Because the DNA fragment is often shorter than the sum of the read lengths, the 3' end of the fragment often contains adapter sequence and other "junk" sequence. The adapter sequence may contain cytosine bases, which will be misinterpreted as evidence of methylation \citep{Krueger:2012ks}. Similarly, "fill-in cytosines" are used in the construction of RRBS libraries to repair the ends of DNA fragments after cleavage by MspI; these would also be misinterpreted as evidence of methylation \citep{Krueger:2012ks}. Another source of M-bias is incomplete or uneven bisulfite-conversion (__CITE__).

\subsubsection{Estimating M-bias}\label{sec:estimating_m-bias}

Estimating the M-bias and incorporating the results into the methylation calling can be done using two different strategies:

1. Compute the M-bias from the filtered data, then call methylation events incorporating both the filters and the M-bias. This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling. This is the approach used by `bismark_methylation_extractor`, `methtuple` and `Bis-SNP`.
2. Call methylation events from the filtered data but retain the read-position of each methylation event. Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias. This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the first file containing the methylation calls. This is the approach taken by `BSmooth`.

I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `methtuple`. In theory, the first pass over the `SAM/BAM` could be "skipped" by computing the M-bias simultaneously with sorting the `SAM/BAM` and marking PCR duplicates. For example, using (idealised) Unix pipes:

```
# 'align' is read mapping software
# 'mark_duplicates' is PCR duplicate marking software. It creates 'out.bam'.
# 'mbias' computes mbias. It produces 'mbias.out' and (passes through) 'out.bam'.
align ref.idx r1.fq r2.fq | mark_duplicates | mbias --filters
# call_meth is methylation calling software. It produces 'call_meth.out' based on 'mbias.out' and 'out.bam'
call_meth --filters mbias.out out.bam
```

@@

```{r}
# TODO: Figure of M-bias computed from raw vs. filtered data
```

```{r}
# TODO: Figure showing several M-bias curves with different patterns, e.g. 5' and 3' junk, within-read spikes/troughs, etc.
```

\subsection{Pre-trimming reads destroys the one-to-one relationship between read-position and sequencing cycle}\label{sec:pre-trimming_reads}

Trimming reads prior to alignment (_pre-trimming_), such as using `Trim Galore!` to remove adapter sequence from reads, destroys the one-to-one relationship between the sequencing cycle and the read position. This causes a minor problem when computing M-bias because we no longer know whether the read position is identical to the sequencing cycle. Soft- or hard-clipping reads of their adapter sequence __during__ the alignment avoids this issue, because the clipping information (should be) preserved in the `CIGAR` string[^cigar].

[^cigar]: Many downstream tools, including the current version of `methtuple`, do not properly handle the information in the `CIGAR` string, particularly for soft-clipped reads. The current version of `methtuple` will skip a read (with a warning) that contains a `I` (insertion), `D` (deletion), `S` (soft-clip) or `H` (hard-clip). __FIX THIS__. This is a shortfall of the downstream tools and not of aligner-based clipping _per se_, but is nonetheless an issue in practice. __TODO: Update what changes are made to `methtuple`. The `bwa-meth` and `LAST` alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark`, cannot soft-clip reads and so require that reads are pre-trimmed (__CITE `bwa-meth` pre-print or paper__).__

@@

For example, suppose we performed $100$ bp single-end sequencing and pre-trimmed the first 20 bp of $90\%$ of the reads. Then, read-position $80$ will comprised of $10\%$ sequencing cycle $80$ and $90\%$ sequencing cycle $100$. It is very likely the sequencing cycle $100$ suffers more from M-bias than does cycle $80$, and so this will appear in the M-bias plot as M-bias at read position $80$. When using `bismark_methylation_extractor` this may necessitate that read-positions $80-100$ are ignored in downstream analyses, which would unnecessarily throw away $20\%$ of the data!

This problem cannot be avoided if reads are pre-trimmed because the trimming information is not preserved. \cite{Hansen:2012gr} suggest a separate M-bias plot for each read-length, which will help mitigate the effect of confounding between read-position and sequencing-cycle[^read_trimming]. However, if trimming is performed during the alignment then all the necessary information is retained and the x-axis of M-bias plots would be "sequencing cycle" rather than "read-position".

>>>>

---
title: "Bioinformatics analysis of whole-genome bisulfite-sequencing data"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

@@

This chapter and the next describe the analysis of a whole-genome bisulfite-sequencing dataset, including the many computational and statistical challenges. There are four fundamental steps in the analysis of whole-genome bisulfite-sequencing data:

1. Data quality control checks
2. Read mapping and post-processing of mapped reads
3. Methylation calling
4. Downstream analyses

Steps 1 and 2 will be familiar to anyone who has analysed high-throughput sequencing data, but each requires a twist to work with bisulfite-sequencing data. Step 3 is obviously unique to assays of DNA methylation, but there are similarities to variant calling from DNA-seq. There are a wide variety of analyses that might be performed at step 4 and these have the greatest need for careful statistical analysis.

This chapter focuses on steps 1-3 and concludes with a description of my methylation calling software, `methtuple`, which uniquely can call methylation events at tuples of loci from whole-genome bisulfite-sequencing data. The following chapter addresses step 4.

In this chapter I concentrate on the methylC-seq protocol, which is a directional protocol and the standard whole-genome bisulfite-sequencing assay. All data used in my thesis were generated using this protocol. Other bisulfite-sequencing assays, particularly targeted assays such as RRBS and Methyl-Seq, require some further customised processing.

@@

The `FastQC` software is a very useful tool for performing this first step ([http://www.bioinformatics.babraham.ac.uk/projects/fastqc/](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)). It produces summary graphs of many key measures such as base quality scores, read length distribution and sequence contamination. `FastQC` is a general purpose tool for performing quality control checks of high-throughput sequencing data. This means that some of its output must be interpreted with caution for bisulfite-sequencing data. For example, `FastQC` will report a warning (resp. error) if the relative frequency of the four nucleotides differ by more than $10\%$ (resp. $20\%$). As noted in the `FastQC` documentation, some a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in their sequence composition.

More important is the identification and removal of contaminating sequences. `FastQC` will screen a subset of the reads against a list of known, common contaminants such as adapter sequences. Illumina adapter sequences are ligated to each DNA molecule in the library in order to perform Illumina sequencing. The sequencer can 'read into' the adapter sequence, particularly when using paired-end sequencing of short DNA fragments such as those created in bisulfite-sequencing libraries. This means that some reads are a chimera that contain the sequence of interest (from the sample) and junk (from the adapters). This contamination needs to be removed for two reasons:

1. Reads containing adapter contamination will generally not map to the reference genome, meaning these reads go to waste.
2. If they do map, then this will result incorrect inferences - the 'garbage in, garbage out' maxim.

Using a tool such as `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) or `cutadapt` \cite{Martin:2011va}, the reads can be _trimmed_ to remove these contaminants. Reads might also be trimmed to remove low quality positions, which are common at the 3' end of reads, although this isn't as essential as trimming to remove contaminants.

@@

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are now mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed. Several review articles have tried to summarise and compare the various approaches \cite{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.

@@

1. Methylation-aware mismatch penalties.
2. _In silico_ bisulfite-conversion of reads and reference genomes.

While methylation-aware mappers provide the highest efficiency, these suffer from a bias whereby methylated reads are preferentially mapped over unmethylated reads \cite{Krueger:2012ks}. This biases downstream inference and means that these mappers are generally less popular. I will instead focus on the _in silico_ bisulfite-conversion mappers, called as such because of a key step taken by these in the mapping process.

_In silico_ bisulfite-conversion mappers convert all cytosines to thymines (resp. guanines to adenines) of the forward (resp. reverse) strand from the reference genome. They then take each read and create two _in silico_ bisulfite-converted versions of it[^2_reads]: the CT-read replaces all residual thymines with cytosines and the GA-read replaces all residual guanines with adenines. The CT-read is mapped against the CT-genome and the GA-read is mapped against the GA-genome using a standard mapping tools such as `Bowtie1` \cite{Langmead:2009fv}, `Bowtie2` \cite{Langmead:2012jh} or `bwa`  \cite{Li:2009fi, Li:2010bl}[^non_directional].

@@

[^non_directional]: If the data were generated using a non-directional protocol, then each read of the CT-read and the GA-read are mapped to both of the CT-genome and GA-genome, resulting in four mapping steps per read.

Depending on the exact settings used, the mapper reports the 'best' location of each read with respect to the two reference genomes. It reports the original sequence of the read in the output file so that the methylation status of each position can be inferred by comparing it to the corresponding reference sequence.

_In silico_ bisulfite-conversion mappers avoid the bias inherent in the methylation-aware mappers because all reads, regardless of methylation status, 'look the same' to the mapper. However, they do suffer from a slight loss in mapping efficiency \cite{Krueger:2012ks}.

Table \ref{tab:bs-seq_mappers} list some popular bisulfite-sequencing read mappers, which have been selected to highlight the variety of back-ends used by these software.

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
Name      & Reference                                            & Underlying mapping software \\ \midrule
Bismark   & \cite{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
bwa-meth  & \cite{Pedersen:2014wt}                             & bwa-mem                     \\
BSMAP     & \cite{Xi:2009bd}                                   & SOAP                        \\
Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
\end{tabular}
\caption{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of back-ends used by these software.}
\label{tab:bs-seq_mappers}
\end{table}

Each of these aligners can report the output in the standard Sequence Alignment/Map format (`SAM`) or it's binary equivalent, `BAM` \cite{Li:2009ka}. However, there is no agreed upon standard in the `SAM` specification for encoding the data specific to bisulfite-sequencing, which means that each mapper does this in a slightly different way. This makes it difficult to develop downstream analysis tools.

@@

The above problems are general challenges of read mapping and are not specific to bisulfite-sequencing data, although the reduced complexity of bisulfite-sequencing reads exacerbates these issues. The difficulty of mapping to repetitive regions of the genome is a particularly frustrating one for  bisulfite-sequencing data. Repetitive sequences, such as LINEs and SINEs, are frequently methylated to prevent transcription and so are of interest to researchers studying DNA methylation. The low mapping efficiencies of these regions means that there is often limited data of these elements from bisulfite-sequencing data (__TODO: CHECK IN MY DATA AND REPORT COVERAGE__).

@@

PCR amplification of the input DNA is a common step in creating a library for high-throughput sequencing. PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work. Unfortunately, it can introduce biases into the library that result in some molecules being over-represented or under-represented compared to their true frequency. This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome. These multiple-sequenced fragments are called _PCR duplicates_.

In bisulfite-sequencing data, PCR duplicates containing the a methylation locus can result in a biased estimate of the methylation level at that locus. This is because the sequenced reads do not accurately represent the true methylation levels of the sample.

Generally, there is no way to tell based on sequencing data if a read is truly a PCR duplicate. However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as "PCR duplicates"[^pcr_dup]). Software to identify PCR duplicates include the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \cite{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}. These software use mapped reads from a `SAM/BAM` file as input. There are subtle differences in how each method calls PCR duplicates. As an example, this is how `Picard`'s `MarkDuplicates` function works:

[^pcr_dup]: The distinction between suspected PCR duplicates and true PCR duplicates is rarely made, probably because the phrase is so clunky. Suspected PCR duplicates are almost always referred to as PCR duplicates with the implicit assumption that the reader is aware that these very likely include false positive calls. Consistent with the literature, I will use the term PCR duplicates when I refer to reads identified as PCR duplicates by some software. I will use _true_ PCR duplicates when I need to distinguish the two concepts.

> Essentially what it does (for pairs; single-end data is also handled) is to find the 5' coordinates and mapping orientations of each read pair. When doing this it takes into account all clipping that has taking place as well as any gaps or jumps in the alignment. You can thus think of it as determining "if all the bases from the read were aligned, where would the 5' most base have been aligned". It then matches all read pairs that have identical 5' coordinates and orientations and marks as duplicates all but the "best" pair. "Best" is defined as the read pair having the highest sum of base qualities as bases with Q >= 15 ( [http://broadinstitute.github.io/picard/faq.html](http://broadinstitute.github.io/picard/faq.html)).

Using these software will result in false positive calls because reads may satisfy this (or similar) criteria yet come from separate DNA fragments in the original sample. The false positive rate is a particular problem when a small 'genome' is sequenced at high coverage, where the 'genome' might be a targetted region of a genome such as exons or CpG islands. This can be thought of as an example of the pigeonhole principle, which states that if we have $m$ containers (positions where a read can start) and $n > m$ items (reads), then at least one container must contain more than one item (at least one position must have more than one read starting there).

We could make this mathematically more precise, but it doesn't give us a simple answer to the question, 'should we remove possible PCR duplicates from bisulfite-sequencing data?'. The unsatisfactory answer is, 'it depends'. A rule of thumb is that provided the average or median sequencing coverage of the 'genome' is less than the fragment length[^pcr_fragment_length] then we expect few false positive calls.

[^pcr_fragment_length]: For single-end data, the 'fragment length' in these calculations is the read length.

@@

Ideally, the probability that a base is called methylated should be independent of the sequencing cycle. \citet{Hansen:2011gu} found that this is not the case and that in fact there is considerable bias towards the start (5') and end (3') of reads. They called this _M-bias_.

M-bias can be identified by plotting the read-position methylation level ($rpml$), which is the proportion of reads that are methylated at each read-position, as a function of read-position. These $rpml$ are computed separately for each methylation type and, for paired-end sequencing data, separately for _read 1_ and _read 2_. If there is no M-bias then this plot should be a horizontal line. A 'bend' or 'spike' in this plot is evidence of M-bias. Furthermore, these lines, which indicate the average level of methylation in the sample for that methylation type, should be at the same height for `read_1` and `read_2`, although we see this is often not quite the case. Figure \ref{fig:ADS_M-bias} shows the M-bias plot for the ADS sample from the Lister (2011) dataset.

```{r m-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:ADS_M-bias", fig.cap = "M-bias plot for the ADS sample from the Lister (2011) dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", fig.width = 8}
library(dplyr)
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
read.mbias <- function(file) {

  # Echo some information
  message(paste0("Reading ", file, " ..."))

  # Read the file.
  x <- read.table(file, sep = '\n', as.is = TRUE)

  # Find 'header' lines, which are dispersed throughout the file.
  header_lines <- sapply(X = x, function(xx) {
    grepl(pattern = 'context', x = xx)
    })
    y <- vector(mode = "list", length = sum(header_lines))
    names(y) <- x[header_lines]

    # z indexes the lines containing M-bias values.
    z <- sapply(X = x, function(xx) {
      grep(pattern = "context|^=|position", x= xx, invert = TRUE)
      })
      dz <- diff(z)
      next_context_idx <- c(which(dz > 1), nrow(z))
      colnames_idx <- sapply(X = x, function(xx) {
        grep(pattern = "position", x = xx)
        })[1]
        colnames <- strsplit(x = x[colnames_idx, ], split = "\t")[[1]]

        # Loop over the lines in the file.
        for (i in seq_along(next_context_idx)) {
          # Note the "+4" to skip the intermediate 'header' rows
          start_idx <- ifelse(i == 1, 4, z[next_context_idx[i - 1]] + 4)
          end_idx <- z[next_context_idx[i]]  
          yy <- x[seq(from = start_idx, to = end_idx, by = 1), ]
          y[[i]] <- do.call("rbind", lapply(yy, function(yyy, colnames) {
            tmp <- strsplit(yyy, split = "\t")
            val <- data.frame(as.integer(tmp[[1]][1]),
            as.integer(tmp[[1]][2]),
            as.integer(tmp[[1]][3]),
            as.numeric(tmp[[1]][4]),
            as.integer(tmp[[1]][5]))
            colnames(val) <- colnames
            return(val)
            }, colnames = colnames))
          }
          y <- do.call("rbind", y)
          y$Context <- strtrim(rownames(y), width = 3)

          # Add R1/R2 annotation if paired-end data
          if (grepl(pattern = "R1", x = rownames(y)[1])){
            y$Read <- ifelse(grepl(pattern = "R1", x = rownames(y)), "R1", "R2")
            } else{
              y$Read <- "SE"
            }
            rownames(y) <- NULL

            # Simplify column names
            colnames(y) <- c('read_pos', 'M', 'U', 'perc_M', 'total',
            'methylation_type', 'read')

            # Re-order columns
            y <- y[, c('read', 'read_pos', 'methylation_type', 'M', 'U', 'total',
            'perc_M')]

            # Replace CpG with CG
            y$methylation_type <- ifelse(y$methylation_type == 'CpG', 'CG',
            y$methylation_type)

            # Convert to an MBias object.
            return(y)
          }
#files <- paste0("../M-bias/", list.files("../M-bias/"))
# ADS
files <- "../M-bias/ADS.M-bias.txt"
sample_names <- gsub("../M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "ADS") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: ADS") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))

```

If a sample was processed in multiple batches, then M-bias estimation (and methylation calling) should perhaps be performed separately for each batch and then combined, or in some manner that is 'batch aware'. For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs. Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with.

The strongest source of M-bias in Illumina WGBS data is at the 5' end of _read 2_, which sequences the 3' end of the DNA fragment. Because the DNA fragment is often shorter than the sum of the read lengths, the 3' end of the fragment often contains adapter sequence and other 'junk' sequence. The adapter sequence may contain cytosine bases, which will be misinterpreted as evidence of methylation \citep{Krueger:2012ks}. Similarly, "fill-in cytosines" are used in the construction of RRBS libraries to repair the ends of DNA fragments after cleavage by MspI; these would also be misinterpreted as evidence of methylation \citep{Krueger:2012ks}. Another source of M-bias is incomplete or uneven bisulfite-conversion.

\subsubsection{Computing M-bias}\label{sec:estimating_m-bias}

Computing the M-bias and incorporating the results into the methylation calling can be done using two different strategies:

1. Compute the M-bias from the aligned reads, then call methylation events. The methylation calling should include filters to remove the detected M-bias (along with any other additional filters). This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling. This is the approach used by `bismark_methylation_extractor` \citep{Krueger:2011eb} and `Bis-SNP` \citep{Liu:2012ge}. ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)).
2. Call methylation events but retain the read-position of each methylation event. Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias. This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the first file containing the methylation calls. This is the approach taken by `BSmooth` \citep{Hansen:2012gr}.

I find the first strategy conceptually simpler, and easier to program, and so use it in my methylation calling software, `methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)). In theory, the M-bias could be estimated during the alignment step or during another processing step, such as sorting or marking PCR duplicates, to avoid an additional pass over the `SAM/BAM` file.

@@

\subsubsection{Pre-trimming reads destroys the one-to-one relationship between read-position and sequencing cycle}\label{sec:pre-trimming_reads}

Trimming reads prior to alignment (_pre-trimming_), such as using `Trim Galore!` ([http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/](http://www.bioinformatics.babraham.ac.uk/projects/trim_galore/)) to remove adapter sequence from reads, destroys the one-to-one relationship between the sequencing cycle and the read position. This causes a minor problem when computing M-bias because we no longer know whether the read position is identical to the sequencing cycle. Soft-clipping or hard-clipping reads of their adapter sequence _during_ the alignment avoids this issue, because the clipping information (should be) preserved in the `CIGAR` string[^cigar].

[^cigar]: Not all software properly handles the information in the `CIGAR` string, particularly for soft-clipped reads. This is a shortfall of the downstream tools and not of aligner-based clipping _per se_, but is nonetheless an issue in practice. The `bwa-meth` \citep{Pedersen:2014wt} and `LAST` \citep{Kieibasa:2011do} alignment software both perform well without pre-trimming of reads because they can soft-clip reads on the fly whereas other bisulfite-sequencing aligners, such as `Bismark` \citep{Krueger:2011eb}, cannot soft-clip reads and so require that reads are pre-trimmed.

@@

For example, suppose we performed $100$ bp single-end sequencing and pre-trimmed the first 20 bp of $90\%$ of the reads. Then, read-position $80$ will comprise $10\%$ sequencing cycle $80$ and $90\%$ sequencing cycle $100$. It is very likely that sequencing cycle $100$ suffers more from M-bias than does cycle $80$, and so this will appear in the M-bias plot as M-bias at read position $80$. As most methylation callers exclude all read positions beyond a certain point, e.g., beyond read position 80, the would unnecessarily throw away $20\%$ of the data. `methtuple` is able to exclude specific read-positions, which allows it to deal with 'spikes' in the M-bias plot, like that shown in Figure \ref{fig:E18VA_M-bias}.
